\ \break Automatikus korpuszépítés tulajdonnév-felismerés céljára

### 2013-01-04 paper/MSZNY2013/mszny_slides.tex
nagyjabol kesz

@@

\begin{frame}{Differences from English}
  \bigskip
  \begin{itemize}
  \vitem Language processing tools
    \begin{itemize}
    \vitem Inflectional morphology: different set of tools were needed 
    \vitem We developed a flexible framework that enables the plugging of any NLP tool
    % TODO: what kind of linguistic tools are needed: table?
    \end{itemize}
  \vitem Wikipedia size and coverage
    \begin{itemize}
    \vitem Hungarian is medium-size language: \nagytilde 200k articles vs.~\nagytilde 4M English pages
    \vitem Cross-references may point to English pages
    \vitem Templates are in Hungarian: per-language template configuration
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Differences from English -- cont.}
  \bigskip
  \begin{itemize}
  \vitem DBpedia coverage
    \smallskip
    \begin{itemize}
    \vitem Internationalized chapters for big languages only
    \vitem "Hungarian" DBpedia list contains English articles % Nem egyértelmű
      \begin{itemize}
      \vitem Hungarian equivalents must be added to the list % Added: lásd előző oldal 2/2
      \vitem Hungarian entities without an English page are missing:
             the NER language model might not be accurate
      \vitem Future work: extend the list with the missing Hungarian entities
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Differences from English -- cont.}
  \bigskip
  \begin{itemize}
  \vitem NE conventions
    \smallskip
    \begin{itemize}
    \vitem Szeged NER guideline: similar to CoNLL
    \vitem Common nouns in Hungarian: names of languages, religions, months, etc. \nyil more consistent MISC category
    \vitem NEs can be subject of compounding \nyil the joint common noun can modify the original sense and category of the NE:
    \end{itemize}
  \end{itemize}
  \bigskip
  \begin{Examplee}
  \small{WorldCom {\normalfont(\texttt{ORG})} \nyil WorldCom-botrány ['WorldCom scandal'] {\normalfont(\texttt{O})}}
  \end{Examplee}
\end{frame}

@@

\begin{frame}{Silver Standard Corpora}

@@

Silver standard corpora can be useful for improving NER in several ways:

\begin{itemize}
\vitem for less resourced languages they can serve as training corpora in lieu of gold standard datasets
\vitem they can serve as training sets for domains differing from newswire
\vitem they can be sources of huge entity lists
\vitem they can be sources of feature extraction
\end{itemize}

@@

\begin{frame}{Evaluation}

@@

A maxent NE tagger was used with the feature set consisting of

\begin{itemize}
\vitem gazetteer features
\vitem sentence start and end position
\vitem Boolean-valued orthographic properties of the word form
\vitem string-valued properties of the word form
\vitem morphological information
\end{itemize}

@@

The CoNLL standard method was used for evaluation.

@@

\begin{frame}{Corpus Size and NE Density}

English:

\begin{center}
\begin{tabular}{llll}
\hline  & \bf enwiki & \bf enwiki filtered & \bf CoNLL \\ \hline
token & 60,520,819 & 21,718,854 & 302,811 \\
NE & 3,169,863 & 3,169,863 & 50,758 \\
NE density & 5.23\% & 14.59\% & 16.76\% \\ \hline
\end{tabular}
\end{center}

Hungarian:

\begin{center}
\begin{tabular}{llll}
\hline  & \bf huwiki & \bf huwiki filtered  & \bf Szeged NER \\ \hline
token & 19,108,027 & 3,512,249  & 225,963 \\
NE & 456,281 & 456,281  & 25,896 \\
NE density & 2.38\% & 12.99\%  & 11.46\% \\ \hline 
\end{tabular}
\end{center}

Quite low NE density \nyil sentences without NEs were filtered out from the evaluation datasets

@@

\begin{frame}{Experiments}

\begin{nix}
Our experiments were performed with a sample of 3.5 million tokens
\end{nix}

@@

\begin{itemize}
\item \textit{CoNLL} train -- \textit{CoNLL} test
\item \textit{enwiki} train -- \textit{enwiki} test
\item \textit{enwiki} train -- \textit{CoNLL} test
\item \textit{CoNLL} train with \textit{wikilists} (gazetteer lists extracted from WP) -- \textit{CoNLL} test
\item \textit{CoNLL} train with \textit{wikitags} (labeling the CoNLL datasets by the model trained on WP, and giving these labels as extra features) -- \textit{CoNLL} test
\end{itemize}

\begin{frame}{English Results}

\begin{center}
\begin{tabular}{lllll}
\hline \bf Train & \bf Test & \bf P (\%) & \bf R (\%) & \bf F (\%) \\ \hline
CoNLL & CoNLL & 85.13 & 85.13 & 85.13 \\
enwiki & enwiki & 72.46 & 73.33 &  72.89 \\
enwiki & CoNLL & 56.55 & 49.77 & 52.94 \\
CoNLL with wikilists & CoNLL & 86.33 & 86.35 & 86.34 \\
CoNLL with wikitags & CoNLL & 85.88 & 85.94 & 85.91 \\
\hline
\end{tabular}
\end{center}

\begin{itemize}
\item training and testing across different corpora decreases F-measure
\item the NE tagger trained on WP does not achieve as high performance tested against CoNLL test set as one trained on its own train set
\item gazetteer lists extracted from WP and training with extra features given by the model trained on WP improve F-measure
\end{itemize}

\begin{frame}{Hungarian Results}

\begin{center}
\begin{tabular}{lllll}
\hline \bf Train & \bf Test & \bf P (\%) & \bf R (\%) & \bf F (\%) \\ \hline
Szeged  & Szeged  & 94.50 & 94.35 & 94.43 \\
huwiki & huwiki & 90.64 & 88.91 &  \textbf{89.76} \\
huwiki & Szeged  & 63.08 & 70.46 & 66.57 \\
Szeged  with wikilists & Szeged  & 95.48 & 95.48 & 95.48 \\
Szeged  with wikitags & Szeged  & 95.38 & 94.92 & 95.15 \\
\hline
\end{tabular}
\end{center}

\begin{itemize}
\vitem F-measures for Hungarian are higher 
\begin{itemize}
\vitem the MISC category is more consistent
\vitem the Szeged NER corpus is highly accurately tagged 
\end{itemize}
\vitem the Hungarian corpus can serve as a training corpus to build NE taggers for non-newswire domains
\end{itemize}

@@

% TODO: nem fér ki minden oszlop
\begin{tabular}{lllllll}
\textbf{token} & \textbf{text/link} & \textbf{WP link} & \textbf{POS} & \textbf{lemma} & \textbf{NE} \\
Neumann & B-link & Neumann János & NOUN & neumann & B-PER & B-PER \\
János & I-link & Neumann János & NOUN & jános & I-PER & I-PER \\
( & text & 0 & PUNCT & ( & O & O \\
1903–1957 & text & 0 & NUM & 1903–1957 & O & O \\
) & text & 0 & PUNCT & ) & O & O \\
matematikai & text & 0 & ADJ & matematikai & O & O \\
szemszögből & text & 0 & NOUN<CAS<ELA{>}> & szemszög & O & O \\
közelíti & text & 0 & VERB<DEF> & közelít & O & O \\
meg & text & 0 & PREV & meg & O & O \\
a & text & 0 & ART & a & O & O \\
kérdést & text & 0 & NOUN<CAS<ACC{>}> & kérdés & O & O \\
\end{tabular}

@@

\section{Összegzés}
\begin{frame}{Conclusions}
  \begin{itemize}
  \vitem We presented freely available NE tagged corpora for English and Hungarian, fully automatically generated from WP
  \vitem We applied a new approach: mapping DBpedia ontology classes to standard CoNLL NE tags, and assigning them to WP entities
  \vitem The huge amount of WP articles opens the possibility of building large enough corpora for less resourced languages such as Hungarian
  \vitem Silver standard corpora can improve NER accuracy in more ways
  \vitem The Hungarian corpus can serve as a training corpus to build NE taggers for non-newswire domains
  \end{itemize}
  \vfill
\end{frame}

>>>>

\begin{frame}{Kézi ellenőrzés}

A teljes korpusz 0,001\%-át kézzel annotáltuk, majd összehasonlítottuk a gépi és kézi annotálás eredményét.

\begin{tabular}{lcccc}
\toprule
& P (\%) &  R (\%) & F$_{\beta=1}$ (\%) & NE (\#) \\
\midrule
LOC     &   98.72 &  95.65 &  97.16 & 161 \\
MISC    &   95.24 &  76.92 &  85.11 & 26 \\
ORG     &   89.66 &  89.66 &  89.66 & 29 \\
PER     &   88.30 &  89.25 &  88.77 & 93 \\
\midrule
Összesítve &   94.33 &  91.59 &  92.94 & 309 \\
\bottomrule
\end{tabular}

\begin{frame}{Igazságmátrix}

\begin{tabular}{l@{\hspace{0.5em}}|@{\hspace{0.5em}}cccc}
\toprule
Auto$\downarrow$ / Gold$\rightarrow$ & PER & ORG & LOC & MISC \\
\midrule
PER & 83 & 1 & & 2 \\
ORG &  & 26 & 1 & 1 \\
LOC &  & 1 & 154 & \\
MISC &  &  & 1 & 20 \\
\bottomrule
\end{tabular}

A típustévesztés okai:

\begin{itemize}
\item a DBpedia típusinformációja helytelen: \\ $[\mbox{MTA}]_{LOC}$
\item a Wikipédia rossz kereszthivatkozásai: \\ $[\mbox{Walt Disney}]_{PER}$ Co.
\end{itemize}

\begin{frame}{További hibák}

\begin{tabular}{l@{\hspace{0.5em}}cccc}
\toprule
	& PER & ORG & LOC & MISC \\
\midrule
Álpozitív & 1 &  0  &  1  &  0 \\
Álnegatív & 3 &  0  &  5  &  4 \\
Részlegesen felismert nevek & 7 & 1 & 0 & 0 \\
\bottomrule
\end{tabular}

\begin{itemize}
\item Álpozitív: hibásan névként felismert szavak: $[\mbox{Bizánc-ellenes}]_{LOC}$
\item Álnegatív: fel nem ismert nevek: $[\mbox{ókori Róma}]_O$
\item Részlegesen felismert nevek: pontatlan entitáshatárok: $[\mbox{Szent István király}]_{PER}$
\end{itemize}

@@

% Dávid
\begin{frame}{Differences from English}
  \bigskip
  \begin{itemize}
  \vitem Language processing tools
    \begin{itemize}
    \vitem Inflectional morphology: different set of tools were needed 
    \vitem We developed a flexible framework that enables the plugging of any NLP tool
    % TODO: what kind of linguistic tools are needed: table?
    \end{itemize}
  \vitem Wikipedia size and coverage
    \begin{itemize}
    \vitem Hungarian is medium-size language: \nagytilde 200k articles vs.~\nagytilde 4M English pages
    \vitem Cross-references may point to English pages
    \vitem Templates are in Hungarian: per-language template configuration
    \end{itemize}
  \end{itemize}
\end{frame}

% Dávid
\begin{frame}{Differences from English -- cont.}
  \bigskip
  \begin{itemize}
  \vitem DBpedia coverage
    \smallskip
    \begin{itemize}
    \vitem Internationalized chapters for big languages only
    \vitem "Hungarian" DBpedia list contains English articles % Nem egyértelmű
      \begin{itemize}
      \vitem Hungarian equivalents must be added to the list % Added: lásd előző oldal 2/2
      \vitem Hungarian entities without an English page are missing:
             the NER language model might not be accurate
      \vitem Future work: extend the list with the missing Hungarian entities
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

% Eszter
% TODO a következő fejezetbe
\begin{frame}{Differences from English -- cont.}
  \bigskip
  \begin{itemize}
  \vitem NE conventions
    \smallskip
    \begin{itemize}
    \vitem Szeged NER guideline: similar to CoNLL
    \vitem Common nouns in Hungarian: names of languages, religions, months, etc. \nyil more consistent MISC category
    \vitem NEs can be subject of compounding \nyil the joint common noun can modify the original sense and category of the NE:
    \end{itemize}
  \end{itemize}
  \bigskip
  \begin{Examplee}
  \small{WorldCom {\normalfont(\texttt{ORG})} \nyil WorldCom-botrány ['WorldCom scandal'] {\normalfont(\texttt{O})}}
  \end{Examplee}
\end{frame}

@@

%\begin{frame}{Silver standard korpuszok}

%\bigskip

%A silver standard korpuszok

%\begin{itemize}
%\vitem for less resourced languages they can serve as training corpora in lieu of gold standard datasets
%\vitem they can serve as training sets for domains differing from newswire
%\vitem they can be sources of huge entity lists
%\vitem they can be sources of feature extraction
%\end{itemize}

\begin{frame}{Kiértékelés}

@@

\begin{itemize}
\item A kiértékeléshez a \texttt{hunner} taggert használtuk az alábbi jegyekkel:
\begin{itemize}
\vitem mondatkezdő- és vég pozíciók
\vitem szóalakon alapuló jegyek
\vitem morfológiai információ
\vitem listajegyek
\end{itemize}
\item Az eredmények kiszámításához a sztenderd CoNLL-módszert alkalmaztuk
\end{itemize}

@@

\begin{frame}{Korpuszméret és telítettség}

@@

\begin{tabular}{lccc}
\toprule
 & \textbf{huwiki} & \textbf{sűrített huwiki} & \textbf{Szeged NER} \\
\midrule
token & 19.108.027 & 3.512.249  & 225.963 \\
NE & 456.281 & 456.281  & 25.896 \\
telítettség (\%) & 2,38 & 12,99 & 11,46 \\
\bottomrule
\end{tabular}

@@

\begin{nix}
szigorú módszer \nyil kevés, de pontos NE \nyil híg korpusz
\end{nix}

A kiértékeléshez eltávolítottuk a NE-t nem tartalmazó mondatokat.

@@

\begin{frame}{Kísérletek}

\begin{nix}
a 3,5 millió tokenes korpuszt 90-10\% arányban tanító és kiértékelő halmazra osztottuk
\end{nix}

\begin{itemize}
\item Szeged NER tanítókorpusz -- Szeged NER tesztkorpusz
\item huwiki tanítóhalmaz -- huwiki teszthalmaz
\item huwiki tanítóhalmaz -- Szeged NER tesztkorpusz
\item Szeged NER tanítókorpusz, a Wikipédiából kinyert névlistákkal -- Szeged NER tesztkorpusz
\item Szeged NER tanítókorpusz, a Wikipédia korpuszon tanított modell által felcímkézve, és a címkék extra jegyként hozzáadva -- Szeged NER tesztkorpusz
\end{itemize}

@@

\begin{frame}{Eredmények}

\begin{center}
\begin{tabular}{lcccc}
\toprule 
\bf tanítás & \bf teszt & \bf P (\%) & \bf R (\%) & \bf F (\%) \\ 
\midrule
Szeged & Szeged & 94,50 & 94,35 & 94,43 \\
huwiki & huwiki & 90,64 & 88,91 &  \textbf{89,76}  \\
huwiki & Szeged & 63,08 & 70,46 & 66,57  \\
Szeged\_wikilisták & Szeged & 95,48 & 95,48 & 95,48  \\
Szeged\_wikitag & Szeged & 95,38 & 94,92 & 95,15 \\
\bottomrule
\end{tabular}
\end{center}

@@

A saját teszthalmazon elért eredmény arról tanúskodik, hogy a korpusz akár önálló gold standard adathalmazként is használható.

@@

% TODO: nem fér ki minden oszlop
%SE: kicsit összébb húztam, de így se fér ki. vagy ki kéne szedni belőle a linkes oszlopokat, vagy valahogy kicsinyíteni az egészet
\begin{tabular}{@{}l@{ }l@{ }l@{ }l@{ }l@{ }l@{ }l}
\textbf{token} & \textbf{text/link} & \textbf{WP link} & \textbf{POS} & \textbf{lemma} & \textbf{NE} \\
Neumann & B-link & Neumann János & NOUN & neumann & B-PER & B-PER \\
János & I-link & Neumann János & NOUN & jános & I-PER & I-PER \\
( & text & 0 & PUNCT & ( & O & O \\
1903–1957 & text & 0 & NUM & 1903–1957 & O & O \\
) & text & 0 & PUNCT & ) & O & O \\
matematikai & text & 0 & ADJ & matematikai & O & O \\
szemszögből & text & 0 & NOUN<CAS<ELA{>}> & szemszög & O & O \\
közelíti & text & 0 & VERB<DEF> & közelít & O & O \\
meg & text & 0 & PREV & meg & O & O \\
a & text & 0 & ART & a & O & O \\
kérdést & text & 0 & NOUN<CAS<ACC{>}> & kérdés & O & O \\
\end{tabular}

@@

\section{Összegzés}
\begin{frame}{Összegzés}
  \begin{itemize}
  \vitem módszerünk nagyban csökkenti az annotációs költségeket 
  \vitem bármely Wikipédia-nyelvre alkalmazható \nyil kevés erőforrással rendelkező nyelvekre is előállítható tulajdonnév-annotált korpusz
  \vitem a létrehozott korpuszok több módon is felhasználhatóak a NER rendszerek hatékonyságának növelésére
  \vitem mivel korpuszunk közelíti a gold standard minőséget, akár önálló tanítóhalmazként is használható a sajtó stílustól eltérő szövegekhez
  \vitem az általunk előállított korpuszok szabadon felhasználhatóak
  \vitem a módszerben benne rejlik a lehetőség finomabb tulajdonnév-hierarchiák támogatására is
  \end{itemize}
  \vfill
\end{frame}
